|

### 2023-05-09 paper/00-main.Rmd
figures + tables formatting
more work

bibliography: bibliography.bib
output: rticles::asa_article
header-includes:
  \renewcommand{\arraystretch}{0.9}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
library(here)
```

>>>>

bibliography: bibliography.bib
biblio-style: chicago
output: rticles::asa_article
# header-includes:
#   \renewcommand{\arraystretch}{0.8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  cache = TRUE
)
library(here)
```

### 2023-05-09 paper/01-introduction.Rmd
more work

Officially tracked by the NFL since 1982, the most well known statistic for evaluating defensive linemen on pass rushing plays is the sack, which occurs when a defensive player tackles the quarterback behind the line of scrimmage before the quarterback is able to throw a pass.  Sacks represent a simple binary measure of success for a rusher on a pass play.  However, there are many intermediate outcomes for a rusher that can also be considered successes that don't result in a sack.  Measures such a hits and hurries are collected by various outlets where hits counts the number of times a QB is hit while passing and hurries counts how often a QB has to throw the ball earlier than they would have liked to.

The sum of sacks, hits, and hurries has been defined as "pressures" (https://www.pro-football-reference.com/years/2022/advanced.htm) and is an improvement over sacks, but suffers from problems of subjectivity (e.g. was the play actually a hurry?) and are still binary outcomes (i.e. There is a continuum of plays that aren't hurries that all counted as simply "not hurries").

Pro football focus (PFF) defines pass-rush productivity as 
$$[ \ (\text{Hurries} + \text{Hits})/2 + \text{Sacks} \ ]\  / \text{ Pass Rush Snaps}$$
This metric gives twice as much weight to a sack relative to hurries and hits, which is a slight improvement, but the choice of weights is ad-hoc an still only measures binary outcomes similar to the shortcomings of previously discussed metrics.   (https://www.pro-football-reference.com/about/advanced_stats.htm)

More recently metrics such as Time In Pocket (TIP) and Pass Rush Win Rate [@Burke2018Created] (PRWR) have also been proposed and are substantial improvements over the less sophisticated counting statistics.  However, these are still imperfect measures as TIP is highly context dependent (e.g. deep vs shallow passing routes).  PRWR is based on whether a pass rusher is able to beat their block before a fixed time (ESPN chooses 2.5 seconds) from the snap, but this is highly dependent on the rather arbitrary choice of time used to define a "win".  In addition, once a cut-off is chosen, this metric becomes binary like most of the previously discussed measures.

(https://www.espn.com/nfl/story/_/id/34536376/2022-nfl-pass-rushing-run-stopping-blocking-leaderboard-win-rate-rankings-top-players-teams)

All of these measures represent different versions of "the best we can do with the available data".  Without more sophisticated data (i.e. player tracking data), it is difficult to measure effectiveness of a pass rusher so many simple metrics of success have been proposed.  Table \ref{tab:metrics} gives a summary of existing pass-rush metrics.

\begin{table}
\caption{A summary of previously-existing pass rush metrics. \label{tab:metrics}}
\centering
\begin{tabular}{p{0.27\textwidth}p{0.67\textwidth}}
  \hline
Metric & Description \\ 
  \hline
Sacks & A defender tackles the QB behind the line of scrimmage before the QB is able to throw a pass \\
Hits & A defender tackles the QB behind the line of scrimmage after the QB is able to throw a pass. \\
Hurries & A defender pressures the QB behind the line of scrimmage such that the QB must release the ball prior sooner than they wanted to. \\
Pressures & $\text{Hurries} + \text{ Hits } + \text{ Sacks}$ \\
Pass-Rush Productivity & $[\ (\text{Hurries} + \text{Hits})/2 + \text{Sacks}\ ]\  / \text{ Pass Rush Snaps}$\\
Time In Pocket & Time (in seconds) between ball snap and throw or pocket collapse for a QB \\
Pass Rush Win Rate & Rate at which pass rusher beats pass block within 2.5 seconds after ball snap\\
   \hline
\end{tabular}
\end{table}

[Pro football ref has different QB pressures formula? https://www.pro-football-reference.com/about/advanced_stats.htm]

@@

The literature on the performance of linemen, either offensive or defensive in football is scant. @AlamarGould2008 studied the the relationship between pass completion rate and performance of offensive linemen in terms of how successful they were at blocking.  The data for this study was collecting essentially by hand by watching and charting the performance of each lineman on a play.  This was done using games from the beginning 2007 season for seven teams and their first three games each.  The authors note that there is not much research on linemen saying "One of the principal reasons for this lack of study is the absence of publicly available data that is needed in order to track the performance of a specific lineman."

Later, @AlamarGoldner2011 again studied offensive linemen performance.  They analyzed 20 teams during the 2010 season for a total of 1,259 plays.  They don't estimate the performance of an individual lineman, but rather positions on a team (e.g. Chicago Bears' left tackle, Pittsburgh Steelers' center, etc.).  They then go on to use a survival analysis model to estimate time in pocket.  The average probability of success from the survival model is then used as a covariate in a model to estimate the probability of competing a pass, and they show that there is a significant relationship between these two probabilities.

>>>>

Table \ref{tab:metrics} gives a summary of existing pass-rush metrics.

Officially tracked by the NFL since 1982, the most well-known statistic for evaluating defensive linemen on pass rushing plays is the sack, which occurs when a defensive player tackles the quarterback behind the line of scrimmage before the quarterback is able to throw a pass.  Sacks represent a simple binary measure of success for a rusher on a pass play.  However, there are many intermediate outcomes for a rusher that can also be considered successes that don't result in a sack.  Measures such a hits and hurries are collected by various outlets where hits counts the number of times a QB is hit while passing and hurries counts how often a QB has to throw the ball earlier than they would have liked to.

<!-- (https://www.pro-football-reference.com/years/2022/advanced.htm) -->

[Pro football ref has a slightly different QB pressures formula than the formula below https://www.pro-football-reference.com/about/advanced_stats.htm]

The sum of sacks, hits, and hurries has been defined as "pressures" and is an improvement over sacks, but suffers from problems of subjectivity (e.g. was the play actually a hurry?) and are still binary outcomes (i.e. There is a continuum of plays that aren't hurries that all counted as simply "not hurries").

<!-- $$\frac{(\text{Hurries} + \text{Hits})/2 + \text{Sacks}}{\text{ Pass Rush Snaps}}$$ -->

Pro football focus (PFF) comes up with a metric called "pass-rush productivity" (PRP), which is a minor modification from the aforementioned pressures metric (see Table \ref{tab:metrics}).
In particular, PRP gives twice as much weight to a sack relative to hurries and hits, which is a slight improvement to pressures.
However, the choice of weights is ad-hoc and still only measures binary outcomes similar to the shortcomings of previously discussed metrics.

<!-- (https://www.pro-football-reference.com/about/advanced_stats.htm) -->

More recently metrics such as Time In Pocket (TIP) and Pass Rush Win Rate (PRWR) [@Burke2018Created] have also been proposed and are substantial improvements over the less sophisticated counting statistics.
Nevertheless, these are still imperfect measures.
TIP is highly context dependent, as it can be influenced by the type of passing route (e.g. deep versus shallow).
PRWR is based on whether a pass rusher is able to beat their block before a fixed time (ESPN chooses 2.5 seconds) from the snap, but this is highly dependent on the rather arbitrary choice of time used to define a "win".
In addition, once a cut-off is chosen, this metric becomes binary like most of the previously discussed measures.

All of these measures represent different versions of "the best we can do with the available data".  Without granular data, it is difficult to measure effectiveness of a pass rusher so many simple metrics of success have been proposed.

\begin{table}
\caption{A summary of previously-existing pass rush metrics. \label{tab:metrics}}
\centering
\begin{tabular}{p{0.27\textwidth}p{0.67\textwidth}}
\hline
Metric & Description \\ 
\hline
Sacks & A defender tackles the QB behind the line of scrimmage before the QB is able to throw a pass \\
Hits & A defender tackles the QB behind the line of scrimmage after the QB is able to throw a pass. \\
Hurries & A defender pressures the QB behind the line of scrimmage such that the QB must release the ball prior sooner than they wanted to. \\
Pressures & $\text{Hurries} + \text{ Hits } + \text{ Sacks}$ \\
Pass-Rush Productivity & $\displaystyle \frac{(\text{Hurries} + \text{Hits})/2 + \text{Sacks}}{\text{ Pass Rush Snaps}}$\\
Time In Pocket & Time (in seconds) between ball snap and throw or pocket collapse for a QB \\
Pass Rush Win Rate & Rate at which pass rusher beats pass block within 2.5 seconds after ball snap\\
\hline
\end{tabular}
\end{table}

@@

The literature on the performance of linemen, either offensive or defensive in football is scant.
@AlamarGould2008 examine the the relationship between pass completion and offensive linemen's performance, in terms of how successful they were at blocking.
The data for this study are manually collected for the first three weeks of the 2007 NFL season, recording whether a lineman holds a block and the time in pocket for the quarterback.

Later, @AlamarGoldner2011 use manually-tracked data during the 2010 season to estimate lineman performance for different team-positions instead of individual defenders (e.g. Chicago Bears' left tackle, Pittsburgh Steelers' center).
[this next part needs re-wording]
They then go on to use a survival analysis model to estimate time in pocket.  The average probability of success from the survival model is then used as a covariate in a model to estimate the probability of competing a pass, and they show that there is a significant relationship between these two probabilities.

### 2023-05-09 paper/03-methods.Rmd
more work

While average STRAIN appears to be positively correlated with pass rusher ability, the observed average STRAIN on a single play is likely due to a convolution of factors. Outside of the pass rusher's ability, there is variability in the opposing strength of pass blockers across plays a pass rusher is involved. Additionally, based on previous work by PFF, the opposing QB plays an important role on observed pressure in passing plays [@eager2019pressure]. Thus, we need to appropriately divide the credit of an observed average STRAIN across the different players involved and other factors.

We use a multilevel model (i.e., hierarchical, random-effects, or mixed effects model) to evaluate pass rushers' impact on the average STRAIN observed in a play, while accounting for their pass blocker, QB, and play-level information. We identify the pass blocker linked with the pass rusher of interest using the data provided by PFF. Since there can be multiple blockers linked to a rusher, for simplicity we only consider the nearest blocker positioned to the pass rusher at the start of the play. We use random intercepts for the three player groups: (1) pass rushers as $R$, (2) nearest pass blockers as $B$, and (3) QBs as $Q$. The coefficients $\beta$ for additional information $X_{j,i}$ about pass rusher $j$ in play $i$ are estimated as fixed-effects. Our model for the average STRAIN by pass rusher $j$ on play $i$ is as follows:

$$
\begin{aligned}
\overline{STRAIN}_{j,i} &\sim Normal(R_{r[j,i]} + B_{b[j,i]} + Q_{q[i]} + X_{j,i} \cdot \beta, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
R_{r} &\sim Normal(\mu_R, \sigma^2_R), \text{ for } r = 1, \cdots, \text{ \# of pass rushers}, \\
B_{b} &\sim Normal(\mu_B, \sigma^2_B), \text{ for } b = 1, \cdots, \text{ \# of pass blockers}, \\
Q_{q} &\sim Normal(\mu_Q, \sigma^2_Q), \text{ for } q = 1, \cdots, \text{ \# of QBs}.
\end{aligned}
$$

We use a Normal distribution to pull the random intercepts for each player toward their respective position group means. This provides us with average individual effects on STRAIN for each player while applying shrinkage, which is a useful property in this problem's context since we do not observe the same number of plays for each player. With regards to the fixed-effects about pass rusher $j$ in play $i$, we include the length of the play, the pass rusher position (INSERT POSITIONS), the nearest pass blocker's position (INSERT POSITIONS), and the number of pass blockers assigned to the pass rusher according to PFF. We fit the model using penalized likelihood via the `lme4` package in `R` [@bateslme4].

>>>>

<!-- While average STRAIN appears to be positively correlated with pass rusher ability -->

While average STRAIN is a simple first step for evaluating the pass rush, the observed average STRAIN on a single play is likely due to a convolution of factors.
Outside of the pass rusher's ability, there is variability in the opposing strength of pass blockers across plays a pass rusher is involved.
Additionally, based on previous work by PFF, the opposing QB plays an important role on observed pressure in passing plays [@eager2019pressure].
Thus, we need to appropriately divide the credit of an observed average STRAIN across the different players involved and other factors.

To this end, we fit a multilevel model to evaluate pass rushers' impact on the average STRAIN observed in a play, while accounting for their pass blocker, QB, and play-level information.
We identify the pass blocker linked with the pass rusher of interest using the data provided by PFF.
Since there can be multiple blockers linked to a rusher, for simplicity we only consider the nearest blocker positioned to the pass rusher at the start of the play.
We use random intercepts for the three player groups: (1) pass rushers as $R$, (2) nearest pass blockers as $B$, and (3) QBs as $Q$.
The coefficients $\beta$ for additional information $X_{j,i}$ about pass rusher $j$ in play $i$ are estimated as fixed-effects. 
Our model for the average STRAIN by pass rusher $j$ on play $i$ is as follows.
\begin{equation}
\begin{aligned}
\overline{\text{STRAIN}}_{j,i} &\sim N(R_{r[j,i]} + B_{b[j,i]} + Q_{q[i]} + X_{j,i} \cdot \beta, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
R_{r} &\sim N(\mu_R, \sigma^2_R), \text{ for } r = 1, \cdots, \text{ \# of pass rushers}, \\
B_{b} &\sim N(\mu_B, \sigma^2_B), \text{ for } b = 1, \cdots, \text{ \# of pass blockers}, \\
Q_{q} &\sim N(\mu_Q, \sigma^2_Q), \text{ for } q = 1, \cdots, \text{ \# of QBs}.
\end{aligned}
\end{equation}

We consider a normal distribution to obtain the random intercepts for each player toward their respective position group means.
This provides us with average individual effects on STRAIN for each player while applying shrinkage, which is a useful property in this problem's context since we do not observe the same number of plays for each player.
With regards to the fixed-effects about pass rusher $j$ in play $i$, we include the length of the play, the pass rusher position, the nearest pass blocker's position, and the number of pass blockers assigned to the pass rusher based on the PFF data.
Table \ref{tab:pos} shows our categorization of position for the pass rush and pass block roles.
We implement the model using penalized likelihood via the `lme4` package in `R` [@bateslme4; @R2023Language].

\begin{table}
\caption{Position groupings for pass rushers and blockers. \label{tab:pos}}
\centering
\begin{tabular}{ll}
\hline
Role & Position \\ 
\hline
Pass rush & Defensive end \\ 
Pass rush & Defensive tackle \\ 
Pass rush & Nose tackle \\ 
Pass rush & Outside linebacker \\ 
Pass rush & Interior linebacker (middle and inside linebackers) \\ 
Pass rush & Secondary (cornerback, free safety, and strong safety) \\
\hline
Pass block & Center \\ 
Pass block & Guard \\ 
Pass block & Tackle \\ 
Pass block & Other (tight end, running back, fullback, wide receiver) \\ 
\hline
\end{tabular}
\end{table}

```{r, eval=FALSE}
rush <- tibble(
  Role = "Pass rush",
  Position = c("Defensive end", "Defensive tackle", "Nose tackle", "Outside linebacker",
               "Interior linebacker (middle and inside linebackers)", 
               "Secondary (cornerback, free safety, and strong safety)")
)
block <- tibble(
  Role = "Pass block",
  Position = c("Center", "Guard", "Tackle", "Other (tight end, running back, fullback, wide receiver)")
)
rush |> 
  bind_rows(block) |> 
  xtable()
```

### 2023-05-09 paper/04-results.Rmd
more work

To illustrate our proposed metric STRAIN for pass rush evaluation, we use the same play from the Las Vegas Raiders and Denver Broncos game as mentioned in Section \ref{sec:data}.
Figure \ref{fig:fig_field} shows an updated version of Figure \ref{fig:fig_field_raw}, with the point size for each Las Vegas defender corresponding to the estimated STRAIN in the selected frames as the play progresses.
In addition, Figure \ref{fig:fig_field} is accompanied by Figure \ref{fig:fig_crosby_curves}, which is a line graph showing how Crosby's distance from the quarterback, velocity, and STRAIN change continuously throughout the play.

>>>>

To illustrate our proposed metric STRAIN for pass rush evaluation, we use the same play from the Las Vegas Raiders and Denver Broncos game as mentioned in Section \ref{sec:data}.
Figure \ref{fig:fig_field} shows an updated version of Figure \ref{fig:fig_field_raw}, with the point size for each Las Vegas defender corresponding to the estimated STRAIN in the selected frames as the play progresses.
This is accompanied by Figure \ref{fig:fig_crosby_curves}, which is a line graph showing how Crosby's distance from the quarterback, velocity, and STRAIN change continuously throughout the play.
